Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints

### 2014-07-17 Compare_053014_statsmed_revisions.tex
Final edits, still waiting on tables

\runninghead{EE Gabriel and MC Sachs and PB Gilbert}
\title{Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints}
\author{Erin E. Gabriel\affil{a}\corrauth\
Michael C. Sachs\affil{b}, Peter B. Gilbert\affil{a,c}}
\address{\affilnum{a}Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center\\
\affilnum{b}Kidney Research Institute, University of Washington\\
\affilnum{c}Department of Biostatistics, University of Washington}
\corraddr{egabriel@fhcrc.org}
\begin{abstract}
Principal surrogate endpoints are useful as targets for Phase I and II trials. In many recent trials, multiple post-randomization biomarkers are measured. However, few statistical methods exist for comparison of or combination of biomarkers as principal surrogates and none of these methods to our knowledge utilize time-to-event clinical endpoint information. We propose a Weibull model extension of the semi-parametric estimated maximum likelihood method of \citet{Huang11} that allows for the inclusion of multiple biomarkers in the same risk model as multivariate candidate principal surrogates. We propose several methods for comparing candidate principal surrogates and evaluating multivariate principal surrogates. These include the time-dependent and surrogate-dependent true and false positive fraction, the time-dependent and the integrated standardized total gain and the cumulative distribution function of the risk difference. We illustrate the operating characteristics of our proposed methods in simulations and outline how these statistics can be used to evaluate and compare candidate principal surrogates. We use these methods to investigate candidate surrogates in the Diabetes Control and Complications Trial.
\end{abstract}
\keywords{Causal inference; Surrogate endpoint evaluation; Survival analysis; Accuracy measures; Multivariate principal stratification}
\maketitle

@@

We propose the use of several curves for displaying PS quality which consider the clinical outcome within levels defined by the marginal causal effect predictiveness (CEP) function rather than the levels of the surrogates themselves. A CEP is a function of the treatment-arm-specific biomarker-conditional risks, such that when the risks are equal the CEP is equal to zero and when the risks are not equal the CEP is not equal to zero. A marginal-CEP is simply a contrast of the clinical outcome over the trial arms within subgroups defined by the candidate surrogate under active treatment; for example, the additive difference in risk between the placebo and treatment groups in those subjects that have or would have had a biomarker measurement. Univariate candidate PS can be evaluated by the strength of the association between the candidate PS under active treatment assignment and the clinical treatment effect. This association can be displayed by the marginal-CEP versus candidate PS curve \citep{Gilbert08}. Evaluation of multivariate principal surrogates is complicated by the increased dimension.

@@

We develop summary statistics for the comparison of candidate PS for a continuous time-to-event endpoint. Most existing methods of evaluating candidate PS are not helpful for ranking candidate PS . Often, evaluation criteria simply provide a way of testing the binary quality of being or not being a PS, rather than providing a quantitative measurement of surrogate quality. We develop two extensions of the standardized total gain \citep{Huang11}, both of which provide a single value measure of surrogate quality that can be used to compare candidate surrogates for the same clinical outcome in the same trial. The time-dependent standardized total gain, ($STG(t)$), allows for the comparison of candidate PS at a given time point $t$ that properly accounts for censoring. Over a range of time points, a display of $STG(t)$ versus $t$ provides information about surrogate quality waning, as well as a means of comparison of candidate surrogates over time. The time integrated  standardized total gain, ($\widetilde{STG}$), which is the time-dependent standardized total gain integrated over the estimated survival times, provides a single value summary of surrogate quality independent of time.

In Section 2, we give some necessary notation and discuss in greater detail why the existing criteria for principal surrogate evaluation are not ideal for the evaluation of multivariate candidate PS or the comparison of candidate PS.  In Section 3.1, we describe the time-dependent risk estimands and outline some assumptions that are helpful for identifying these estimands.  We outline our proposed extension to the estimation methods of \citet{Huang11} and \citet{Gabriel13} in Section 3.2.  We describe our proposed evaluation methods in Sections \ref{CAM}-\ref{CDF} and in Sections \ref{STG}-\ref{ISTG} we detail our proposed summary statistics for comparison. In Section 3.7, we outline our suggested procedure for biomarker comparison using the methods developed in Sections \ref{CAM}-\ref{ISTG}. In Section 4 we evaluate our methods in a simulation study of both univariate and multivariate candidate PS. In Section 5, we apply our methods to the Diabetes Control and Complications Trial (DCCT). We find evidence to support the high quality surrogate identified in \citet{Gabriel13} is better than an alternative candidate surrogate and that combinations of the two candidates are not improvements over the higher quality surrogate alone. In Section 6, we discuss potential limitations of our methods and future avenues for research.

@@

In \citet{Gilbert08} and most of the recent literature, marginal risk estimands are used instead of joint risk estimands because they are easier to identify and estimate. marginal risk estimands condition on $S(1)$ and marginalize over the distribution of $S(0)$. The concept of wide variability is easily extended to marginal risk, such that contrasts of risk over the arms of the trial, such as our CEP of risk difference, vary widely in $s_1$. This can simply be viewed as a subgroup analysis, where subgroups defined by levels of $S(1)$, should strongly modify the clinical treatment effect. Other than the special case when the candidate surrogate is constant in the control arm, $S(0)=C$ called Constant Biomarker (CB), ACN cannot be verified based on the marginal risks alone \citep{Gilbert08, Gabriel13}. \citet{Gabriel13} outlines some PS candidate constructions that can make CB more likely to hold in a general clinical trial where biomarkers under control will tend not to be constant.

@@

After a set of candidate multivariate or univariate candidate surrogates is determine to have some value as PS, it is of interest to rank them and test to determine if any one candidate is significantly better than others. ANC can only classify candidates into two groups. Our suggested accuracy measures and the CDF of the risk model CEP can help more finely group candidates by predictive quality and amount of variation in CEP, but they do not provide inference for comparing candidates. We propose the use of the standardized total gain for candidate surrogate ranking and tests of the difference in standardized total gain between two candidates for comparison inference. These are outlined in Sections \ref{STG} and \ref{ISTG}.

@@

Given assumptions A1-A4 we can link the time-dependent potential outcome risk estimands of interest to observed data clinical risk estimands and the partially observed candidate surrogates by assuming a parametric survival model. We will assume a Weibull model, the probability density function (pdf) for which we will denote as $g(\cdot|\cdot,\cdot)$. We parameterize the scale term of this model, $\gamma=(\gamma_{00}, \gamma_{10},\gamma_{0j}, \gamma_{1j})$, and allow for a constant shape term, $\beta_{0}$; this parameterization is equivalent to a classical Weibull survival model. The density is given by:
\begin{eqnarray*}
g(t|\gamma, \beta, z, s_{1,1}, \ldots,s_{1,J}, y)=\lambda_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J})^{y} Q_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J}),
\end{eqnarray*}
where $Q_z(t|\gamma, \beta, s_{1,j})$ is the conditional Weibull survivor function for treatment arm $Z=z$ and $\lambda_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J})$ is the conditional Weibull hazard function. We assume the risk estimands are of the form:
\begin{itemize}
\item A5: $risk_{z}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) \equiv 1-\exp(-(t/\exp(\gamma_{z0}+\sum_j{\gamma_{zj}s_{1,j})})^{(\exp(\beta_{0}))}$
\end{itemize}
for $z\in \{0,1\}$. We focus on this classical model in the main text for clarity; more complex time-dependent models are outlined in Appendix B of the supplementary materials. In order for both assumptions A3 and A5 to hold simultaneously, we can only consider censoring or event times greater than $\tau$. This can be accomplished in practice by rescaling all follow-up times from $\tau$ rather than from enrollment.

@@

If the model for risk difference that conditions on $\{s_{1,1}, \ldots, s_{1,J}\}$, indicates there is a high probability of having a risk difference above a given threshold for all subjects that failed on non-active treatment but not on active treatment at or before time $t$, this is evidence of the predictive power of the model. The summary $TPF(t|c)$ is therefore a good measure for the comparison or evaluation of candidate PS. Similarly, $\mbox{Specificity}(t|c)$ or $1-FPF(t|c)$, is of interest as those subjects not helped by treatment should have a high probability of having a small risk difference based on a predictive model. As these measures are threshold dependent, they can be used to compare not just candidate surrogates but thresholds for a given surrogate. Thresholds $c$ found to have high $TPF(t|c)$ and low $FPF(t|c)$ are ideal.

It is of interest to determine the relationship between $TPF(t|c)$ and $FPF(t|c)$. The time-dependent ROC curve gives an illustration of the $TPF(t|c)$ and $FPF(t|c)$ relationship at given levels of $c$. This curve is defined by $ROC^{t}(q)=TPF\{t|FPF^{-1}(t|q)\}$ for a given time point $t$, adapting the binary-clinical endpoint $ROC(q)$ definition of \citet{Huang12b}. Figure \ref{4panel} panels B and D displays the $ROC^{t}(q)$ curves for several different surrogate quality levels. The most left-upper point on the $ROC^{t}(q)$ curve is of interest for comparison of candidate surrogates, as candidates with left-upper points closer to $(0,1)$ will tend to be better PS, as elaborated in Section \ref{STG}.
\begin{sidewaysfigure}[H]
\begin{center}
%\includegraphics[scale=1.75, angle=270, clip=TRUE, trim=0 0 20mm 0]{4Panel_trueTE.pdf}
\includegraphics[scale=0.80]{Figure1.jpeg}
\end{center}
\caption{Panel A displays the CDF of the risk difference based on a useless and an ideal univariate surrogate. Panel B displays the ROC curves for the same ideal and useless surrogate candidates. Panel C displays the CDF of risk difference based on a medium quality univariate partial surrogate, and the CDF of risk difference based on a multivariate surrogate, which is the combination of two medium quality univariate partial surrogates. Panel D displays the ROC curves for the same medium quality univariate partial surrogate and multivariate surrogate. \label{4panel}}
\end{sidewaysfigure}

@@

 Let $CDF^{t}_{\Delta}(c)$ be the CDF of the risk difference $\Delta(t|S_{1,1}, \dots, S_{1,J})$, $CDF^{t}_{\Delta}(c)\equiv P\{\Delta(t|S_{1,1}, \dots, S_{1,J})\leq c\}.$ The $CDF^{t}_{\Delta}(c)$ versus $c$ curve is of interest for the evaluation of multivariate candidate PS and for the comparison of candidate PS. An ideal PS will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot that is a step function with a horizontal line at the proportion of subjects with $D(t)=-1$ and a horizontal line at the proportion of subjects with $D(t)=0$; where as defined previously $D(t)=Y^{t}(0)-Y^{t}(1)$. If we assume monotonicity of treatment effect (assumption A6) then the ideal PS candidate will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot with a single horizontal line at the proportion of subjects with $D(t)=0$ (in the case that A6 is $T(0) \leq T(1)$). If we alternatively assume no benefit, then the line will be at the proportion of subjects with $D(t)=-1$. Assumption A6 is not needed for the function $CDF^{t}_{\Delta}(c)$ to be well-defined. For a useless surrogate, the plot of $CDF^{t}_{\Delta}(c)$ versus $c$ will be a vertical line at $\rho_0(t)-\rho_1(t)$, the difference in prevalence of the potential clinical outcomes between the trial arms. Partially useful surrogates will have $CDF^{t}_{\Delta}(c)$ versus $c$ curves between these two curves with better surrogates having flatter curves. Figure \ref{4panel} displays the CDF of the risk difference for several different surrogate quality levels. We suggest the plug-in estimator of $\widehat{CDF}^{t}_{\Delta}(c)$, $\widehat{CDF}^{t}_{\Delta}(c)=\int{\mathrm{I}_{[\hat{\Delta}(t|s_{1,1,i}, \dots, s_{1,j,i})\leq c]} d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})}$. Although we display a useless surrogate in Figure \ref{4panel}, we suggest testing for any surrogate value prior to estimation of $\widehat{CDF}^{t}_{\Delta}(c)$ as it was found in simulations that there is evidence of bias when there is no variation in risk difference; a suggested test is given in Section \ref{CRE}.

@@

The P-value is 0.018 for testing that the estimated $\widetilde{STG}$ difference is zero between HBA1C alone and the univariate linear combination of HBA1C and EGFR. This is marginal evidence that HBA1C alone is superior to the univariate linear combination of HBA1C-EGFR. The test for difference between $\widehat{\widetilde{STG}}$ for HBA1C alone and the multivariate candidate PS provides no evidence of a significant difference, P-value of 0.2.

Figure \ref{exp} Panel A depicts the $\widehat{CDF}^{7.5}(c)$ versus $c$ curves for all four candidate surrogates, HBA1C, EGFR the univariate linear combination HBA1C-EGFR and the multivariate combination. The EGFR curve is most similar to the marginal risk difference line, supporting the finding that EGFR has little value as a PS. The multivariate combination surrogate which is flatter than the EGFR alone curve is very similar to the HBA1C curve. This provides further evidence that the combination candidate does not improve upon HBA1C alone. As there is no evidence of violation of A6, Figure \ref{exp} Panel B displays the $\widehat{ROC}^{7.5}(q)$ versus $q$ curves for the HBA1C and the two combination candidates. Figure \ref{exp} Panel C depicts the $STG(t)$ versus $t$ for $t$ greater than one year for HBA1C and the two combination candidates. Both Panels B and C support that there is no significant difference between HBA1C alone and the multivariate combination, but that HBA1C alone is significantly better than the univariate linear combination candidate.

@@

Although our proposed estimands and estimators for the classification accuracy curves depend on assumption A6 to be well-defined, we show that when there are small undetectable violations of A6, the estimates approximate the desirable modified estimands of no-benefit or unaffected ROC curves (Appendix A supplementary materials). Even when A6 is clearly violated, the suggested $CDF^{t}_{\Delta}(c)$ and $STG(t)$ versus $t$ curves allow for visual comparison of candidate surrogates and the $STG(t)$ or $\widetilde{STG}$ difference tests perform well for discriminating between two candidate PS that both have some value as a PS. While only the risk estimands themselves and the proposed $CDF^{t}(c)$ curve perform well for biomarkers with no surrogate value, our proposed coefficient-based test of any surrogate value has satisfactory power and correct Type 1 error rate. Using the proposed methods we were able to show that the DCCT candidate surrogate HBA1C has more value as a PS than EGFR, and that the combination of HBA1C and EGFR in the same risk model was not an improvement over HBA1C alone.

Our proposed semi-parametric EML estimators seem to perform well when there is a predictive BIP with correlation with the candidate PS of at least 0.5. The performance decreases with the correlation of the BIP and $S_j(1)$, as has been illustrated for previous EML methods. This is a limitation of EML based methods, however, for the motivating example the BIP correlation exceeds $0.5$. We also find that for higher dimensional candidate surrogates vector BIPs can improve convergence properties over univariate BIPs. Although it is not the focus of this work, our model allows for time-variation in the hazard, an extension to the existing time-to-event PS evaluation methods \citep{Qin07, Miao13}. This extension is discussed in greater detail in the supplementary materials where we consider a more complex Weibull model allowing for time-variation in the treatment effect and the surrogate quality following \citet{Gabriel13}.

>>>>

\runninghead{EE Gabriel and MC Sachs and PB Gilbert}
\title{Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints}
\author{Erin E. Gabriel\affil{a}\corrauth\
Michael C. Sachs\affil{b}, Peter B. Gilbert\affil{a,c}}
\address{\affilnum{a}Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center\\
\affilnum{b}Biometric Research Branch, National Cancer Institute\\
\affilnum{c}Department of Biostatistics, University of Washington}
\corraddr{egabriel@fhcrc.org}
\begin{abstract}
Principal surrogate endpoints are useful as targets for Phase I and II trials. In many recent trials, multiple post-randomization biomarkers are measured. However, few statistical methods exist for comparison of or combination of biomarkers as principal surrogates and none of these methods to our knowledge utilize time-to-event clinical endpoint information. We propose a Weibull model extension of the semi-parametric estimated maximum likelihood method of \citet{Huang11} that allows for the inclusion of multiple biomarkers in the same risk model as multivariate candidate principal surrogates. We propose several methods for comparing candidate principal surrogates and evaluating multivariate principal surrogates. These include the time-dependent and surrogate-dependent true and false positive fraction, the time-dependent and the integrated standardized total gain and the cumulative distribution function of the risk difference. We illustrate the operating characteristics of our proposed methods in simulations and outline how these statistics can be used to evaluate and compare candidate principal surrogates. We use these methods to investigate candidate surrogates in the Diabetes Control and Complications Trial.
\end{abstract}
\keywords{Causal inference; Surrogate endpoint evaluation; Survival analysis; Accuracy measures; Multivariate principal stratification}
\maketitle

@@

We propose the use of several curves for displaying PS quality which consider the clinical outcome within levels defined by the marginal causal effect predictiveness (CEP) function rather than the levels of the surrogates themselves. A CEP is a function of the treatment-arm-specific biomarker-conditional risks, such that when the risks are equal the CEP is equal to zero and when the risks are not equal the CEP is not equal to zero. A marginal-CEP is simply a contrast of the clinical outcome over the trial arms within subgroups defined by the candidate surrogate under active treatment; for example, the additive difference in risk between the placebo and treatment groups in those subjects that have or would have had a biomarker measurement of $x$ under treatment. Univariate candidate PS can be evaluated by the strength of the association between the candidate PS under active treatment assignment and the clinical treatment effect. This association can be displayed by the marginal-CEP versus candidate PS curve \citep{Gilbert08}. Evaluation of multivariate principal surrogates is complicated by the increased dimension.

@@

We develop summary statistics for the comparison of candidate PS for a continuous time-to-event endpoint. Most existing methods of evaluating candidate PS are not helpful for ranking candidate PS. Often, evaluation criteria simply provide a way of testing the binary quality of being or not being a PS, rather than providing a quantitative measurement of surrogate quality. We develop two extensions of the standardized total gain \citep{Huang11}, both of which provide a single value measure of surrogate quality that can be used to compare candidate surrogates for the same clinical outcome in the same trial. The time-dependent standardized total gain, ($STG(t)$), allows for the comparison of candidate PS at a given time point $t$ that properly accounts for censoring. Over a range of time points, a display of $STG(t)$ versus $t$ provides information about surrogate quality waning, as well as a means of comparison of candidate surrogates over time. The time integrated  standardized total gain, ($\widetilde{STG}$), which is the time-dependent standardized total gain integrated over the estimated survival times, provides a single value summary of surrogate quality independent of time.

In Section 2, we give some necessary notation and discuss in greater detail why the existing criteria for principal surrogate evaluation are not ideal for the evaluation of multivariate candidate PS or the comparison of candidate PS.  In Section 3.1, we describe the time-dependent risk estimands and outline some assumptions that are helpful for identifying these estimands.  We outline our proposed extension to the estimation methods of \citet{Huang11} and \citet{Gabriel13} in Section 3.2.  We describe our proposed evaluation methods in Sections \ref{CAM}-\ref{CDF} and in Sections \ref{STG}-\ref{ISTG} we detail our proposed summary statistics for comparison. In Section 3.7, we outline our suggested procedure for biomarker comparison using the methods developed in Sections \ref{CAM}-\ref{ISTG}. In Section 4 we evaluate our methods in a simulation study of both univariate and multivariate candidate PS. In Section 5, we apply our methods to the Diabetes Control and Complications Trial (DCCT). We find evidence to support that the high quality surrogate identified in \citet{Gabriel13} is better than an alternative candidate surrogate and that the combination of the two candidates is not an improvement over the higher quality surrogate alone. In Section 6, we discuss potential limitations of our methods and future avenues for research.

@@

In \citet{Gilbert08} and most of the recent literature, marginal risk estimands are used instead of joint risk estimands because they are easier to identify and estimate. Marginal risk estimands condition on $S(1)$ and marginalize over the distribution of $S(0)$. The concept of wide variability is easily extended to marginal risk, such that contrasts of risk over the arms of the trial, such as our CEP of risk difference, vary widely in $s_1$. This can simply be viewed as a subgroup analysis, where subgroups defined by levels of $S(1)$, should strongly modify the clinical treatment effect. Other than the special case when the candidate surrogate is constant in the control arm, $S(0)=C$ called Constant Biomarker (CB), ACN cannot be verified based on the marginal risks alone \citep{Gilbert08, Gabriel13}. \citet{Gabriel13} outlines some PS candidate constructions that can make CB more likely to hold in a general clinical trial where biomarkers under control will tend not to be constant.

@@

After a set of candidate multivariate or univariate candidate surrogates is determine to have some value as PS, it is of interest to rank them and test to determine if any one candidate is significantly better than others. ANC can only classify candidates into two groups. Our suggested accuracy measures and the CDF of the risk model CEP can help more finely group candidates by predictive quality and amount of variation in CEP, but they do not provide inference for comparing candidates. We propose the use of the standardized total gain for candidate surrogate ranking and tests of the difference in standardized total gain between two candidates for inference. These are outlined in Sections \ref{STG} and \ref{ISTG}.

@@

Given assumptions A1-A4 we can link the time-dependent potential outcome risk estimands of interest to observed data clinical risk estimands and the partially observed candidate surrogates by assuming a parametric survival model. We will assume a Weibull model, the probability density function (pdf) which we will denote as $g(\cdot|\cdot,\cdot)$. We parameterize the scale term of this model, $\gamma=(\gamma_{00}, \gamma_{10},\gamma_{0j}, \gamma_{1j})$, and allow for a constant shape term, $\beta_{0}$; this parameterization is equivalent to a classical Weibull survival model. The density is given by:
\begin{eqnarray*}
g(t|\gamma, \beta, z, s_{1,1}, \ldots,s_{1,J}, y)=\lambda_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J})^{y} Q_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J}),
\end{eqnarray*}
where $Q_z(t|\gamma, \beta, s_{1,j})$ is the conditional Weibull survivor function for treatment arm $Z=z$ and $\lambda_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J})$ is the conditional Weibull hazard function. We assume the risk estimands are of the form:
\begin{itemize}
\item A5: $risk_{z}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) \equiv 1-\exp(-(t/\exp(\gamma_{z0}+\sum_j{\gamma_{zj}s_{1,j})})^{(\exp(\beta_{0}))}$
\end{itemize}
for $z\in \{0,1\}$. We focus on this classical model in the main text for clarity; more complex time-dependent models are outlined in Appendix B of the supplementary materials. In order for both assumptions A3 and A5 to hold simultaneously, we can only consider censoring or event times greater than $\tau$. This can be accomplished in practice by rescaling all follow-up times from $\tau$ rather than from enrollment.

@@

If the model for risk difference that conditions on $\{s_{1,1}, \ldots, s_{1,J}\}$, indicates there is a high probability of having a risk difference above a given threshold for all subjects that failed on non-active treatment but not on active treatment at or before time $t$, this is evidence of the predictive power of the model. The summary $TPF(t|c)$ is therefore a good measure for the comparison or evaluation of candidate PS. Similarly, $\mbox{Specificity}(t|c)$ or $1-FPF(t|c)$, is of interest as those subjects not helped by treatment should have a high probability of having a small risk difference based on a predictive model. As these measures are threshold dependent, they can be used to compare not just candidate surrogates but thresholds for a given risk model. Thresholds $c$ found to have high $TPF(t|c)$ and low $FPF(t|c)$ are ideal.

It is of interest to determine the relationship between $TPF(t|c)$ and $FPF(t|c)$. The time-dependent ROC curve gives an illustration of the $TPF(t|c)$ and $FPF(t|c)$ relationship at given levels of the risk difference. This curve is defined by $ROC^{t}(q)=TPF\{t|FPF^{-1}(t|q)\}$ for a given time point $t$, adapting the binary-clinical endpoint $ROC(q)$ definition of \citet{Huang12b}. Figure \ref{4panel} panels B and D displays the $ROC^{t}(q)$ curves for several different surrogate quality levels. The most left-upper point on the $ROC^{t}(q)$ curve is of interest for comparison of candidate surrogates, as candidates with left-upper points closer to $(0,1)$ will tend to be better PS, as elaborated in Section \ref{STG}.
\begin{sidewaysfigure}[H]
\begin{center}
%\includegraphics[scale=1.75, angle=270, clip=TRUE, trim=0 0 20mm 0]{4Panel_trueTE.pdf}
\includegraphics[scale=0.80]{Figure1.jpeg}
\end{center}
\caption{Panel A displays the CDF of the risk difference based on a useless and an ideal univariate surrogate. Panel B displays the ROC curves for the same ideal and useless surrogate candidates. Panel C displays the CDF of risk difference based on a medium quality univariate partial surrogate, and the CDF of risk difference based on a multivariate surrogate, which is the combination of two medium quality univariate partial surrogates. Panel D displays the ROC curves for the same medium quality univariate partial surrogate and multivariate surrogate. \label{4panel}}
\end{sidewaysfigure}

@@

 Let $CDF^{t}_{\Delta}(c)$ be the CDF of the risk difference $\Delta(t|S_{1,1}, \dots, S_{1,J})$, $CDF^{t}_{\Delta}(c)\equiv P\{\Delta(t|S_{1,1}, \dots, S_{1,J})\leq c\}.$ The $CDF^{t}_{\Delta}(c)$ versus $c$ curve is of interest for the evaluation of multivariate candidate PS and for the comparison of candidate PS. An ideal PS will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot that is a step function with a horizontal line at the proportion of subjects with $D(t)=-1$ and a horizontal line at the proportion of subjects with $D(t)=0$; where as defined previously $D(t)=Y^{t}(0)-Y^{t}(1)$. If we assume monotonicity of treatment effect (assumption A6) then the ideal PS candidate will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot with a single horizontal line at the proportion of subjects with $D(t)=0$ (in the case that A6 is $T(0) \leq T(1)$). If we alternatively assume no benefit, then the line will be at the proportion of subjects with $D(t)=-1$. Assumption A6 is not needed for the function $CDF^{t}_{\Delta}(c)$ to be well-defined. For a useless surrogate, the plot of $CDF^{t}_{\Delta}(c)$ versus $c$ will be a vertical line at $\rho_0(t)-\rho_1(t)$, the difference in prevalence of the potential clinical outcomes between the trial arms. Partially useful surrogates will have $CDF^{t}_{\Delta}(c)$ versus $c$ curves between these two curves with better surrogates having flatter curves. Figure \ref{4panel} displays the CDF of the risk difference for several different surrogate quality levels. We suggest the plug-in estimator of $\widehat{CDF}^{t}_{\Delta}(c)$, $\widehat{CDF}^{t}_{\Delta}(c)=\int{\mathrm{I}_{[\hat{\Delta}(t|s_{1,1,i}, \dots, s_{1,j,i})\leq c]} d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})}$.

@@

The P-value is 0.018 for testing that the estimated $\widetilde{STG}$ difference is zero between HBA1C alone and the univariate linear combination of HBA1C and EGFR. This is marginal evidence that HBA1C alone is superior to the univariate linear combination of HBA1C-EGFR. The test for difference between $\widehat{\widetilde{STG}}$ for HBA1C alone and the multivariate candidate PS provides no evidence of a significant difference, P-value of 0.2. Although the multivariate surrogate has a notably higher STG, given the lack of statistical evidence and the increased burden of measuring a second surrogate, we would still favor the use of HBA1C alone.

Figure \ref{exp} Panel A depicts the $\widehat{CDF}^{7.5}(c)$ versus $c$ curves for all four candidate surrogates, HBA1C, EGFR the univariate linear combination HBA1C-EGFR and the multivariate combination. The EGFR curve is most similar to the marginal risk difference line, supporting the finding that EGFR has little value as a PS. The multivariate combination surrogate which is flatter than the EGFR alone curve is very similar to the HBA1C curve. This provides further evidence that the combination candidate does not improve upon HBA1C alone. As there is no evidence of violation of A6, Figure \ref{exp} Panel B displays the $\widehat{ROC}^{7.5}(q)$ versus $q$ curves for the HBA1C and the two combination candidates. Figure \ref{exp} Panel C depicts the $STG(t)$ versus $t$ for $t$ greater than one year for HBA1C and the two combination candidates. Both Panels A and C support that there is no significant difference between HBA1C alone and the multivariate combination, but that HBA1C alone is significantly better than the univariate linear combination candidate. Panel C indicates that there may be some value in the combination surrogate above the HBA1C alone, however due to the high variance, there is insufficient evidence that this difference is significant.

@@

Although our proposed estimands and estimators for the classification accuracy curves depend on assumption A6 to be well-defined, we show that when there are small undetectable violations of A6, the estimates approximate the desirable modified estimands of no-benefit or unaffected ROC curves (Appendix A supplementary materials). Even when A6 is clearly violated, the suggested $CDF^{t}_{\Delta}(c)$ and $STG(t)$ versus $t$ curves allow for visual comparison of candidate surrogates and the $STG(t)$ or $\widetilde{STG}$ difference tests perform well for discriminating between two candidate PS that both have some value as a PS. While only the risk estimands themselves perform well for biomarkers with no surrogate value, our proposed coefficient-based test for discrimination between potentially useful and useless surrogates has good power and correct Type 1 error rate. Using the proposed methods we were able to show that the DCCT candidate surrogate HBA1C has more value as a PS than EGFR, and that the combination candidates were not a significant improvement over HBA1C alone.

Our proposed semi-parametric EML estimators seem to perform well when there is a predictive BIP with correlation with the candidate PS of at least 0.5. The performance decreases with the correlation of the BIP and $S_j(1)$, as has been illustrated for previous EML methods; this is a limitation of EML based methods. We find that for higher dimensional candidate surrogates vector BIPs can improve convergence properties over univariate BIPs for our method. Further investigation of the effect of using a vector BIP on the efficiency for this and other EML based methods is of research interest. Although it is not the focus of this work, our model allows for time-variation in the hazard, an extension to the existing time-to-event PS evaluation methods \citep{Qin07, Miao13}. Further extension of these methods to allow even more flexible models may be useful for some surrogate evaluation settings.
